---
title: "MultiCluster"
author: "J. Meijer"
date: "2025-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```




Check for weird numbers:
- chrfigs_fam_age
- chrchs_timeslept



Check if variables with only one option are removed.


```{r}
# Load libraries
library(flextable)
library(officer)
library(dplyr)
library(readr)
library(readxl)

```

# Create and load basetable

```{r, tidy=TRUE}
#Load function
source("/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Code/GIT/Scripts/Modules/Basetable/Basetable_function.R")

# Set path to dictionary
dictionary_DIR <- "/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Feature selection/Complete_dictionary.xlsx"

#Set path to data
data_DIR <- file.path( "/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Data/72662")

timepoint = 'baseline'

#Run basetable function. 
results <- create_basetable(
  data_DIR = data_DIR,
  dictionary_DIR = dictionary_DIR,
  release = "Amp_scz3",
  timepoint= "baseline"
)

data <- results$data
meta <- results$meta

 
results_screening <- create_basetable(
  data_DIR = data_DIR,
  dictionary_DIR = dictionary_DIR,
  release = "Amp_scz3",
  timepoint= "screening"
)

data_screening <- results_screening$data
meta_screening <- results_screening$meta


results_month2 <- create_basetable(
  data_DIR = data_DIR,
  dictionary_DIR = dictionary_DIR,
  release = "Amp_scz3",
  timepoint= "month_2"
)

data_month2 <- results_month2$data
meta_moth2 <- results_month2$meta

```

```{r}

## 1. Add any columns that exist only in screening ----------------------------
new_cols   <- setdiff(names(data_screening), names(data))
merged_data <- data %>%                     # baseline rows are the backbone
  left_join(select(data_screening, src_subject_id, all_of(new_cols)),
            by = "src_subject_id")

## 2. Decide, variable-by-variable, whether to keep baseline or screening -----
common_vars <- setdiff(intersect(names(data), names(data_screening)),
                       "src_subject_id")    # overlap apart from the key

# Count how many non-missing / non-empty values each data set has
baseline_n  <- sapply(common_vars,
                      function(v) sum(!is.na(data[[v]]) & data[[v]] != ""))
screening_n <- sapply(common_vars,
                      function(v) sum(!is.na(data_screening[[v]]) &
                                      data_screening[[v]] != ""))

vars_take_screening <- names(screening_n)[screening_n > baseline_n]

## 3. Overwrite those columns *row-wise* with screening values ---------------
if (length(vars_take_screening) > 0) {
  merged_data[vars_take_screening] <-
    data_screening %>%
      select(src_subject_id, all_of(vars_take_screening)) %>%
      right_join(select(merged_data, src_subject_id), by = "src_subject_id") %>%
      select(all_of(vars_take_screening))
}

# merged_data now contains:
#   • baseline values everywhere baseline is as complete or more complete
#   • screening values for columns where screening had strictly fewer missing values
```

### Remove useless columns
```{r}
## Remove columns with only 1 unique value
merged_data <- merged_data[, sapply(merged_data, function(col) length(unique(col))>1)]

## Data preprocessing: Traumatic brain injury

# Combine age with yes or no column to get rid of the Non applicable (-3)
if("chrtbi_age_inj" %in% colnames(merged_data)){
  merged_data$chrtbi_age_inj <- paste(merged_data$chrtbi_subject_head_injury, merged_data$chrtbi_subject_age)
  merged_data$chrtbi_age_inj[merged_data$chrtbi_age_inj=="No NA"] <- "No"
  merged_data$chrtbi_age_inj[merged_data$chrtbi_age_inj=="NA NA"] <- NA
}


# Combine how many times with yes or no column to get rid of the Non applicable (-3)
if("chrtbi_ntimes" %in% colnames(merged_data)){
  merged_data$chrtbi_ntimes <- paste(merged_data$chrtbi_subject_head_injury, merged_data$chrtbi_subject_times)
  merged_data$chrtbi_ntimes[merged_data$chrtbi_ntimes=="No NA"] <- "No"
  merged_data$chrtbi_ntimes[merged_data$chrtbi_ntimes=="NA NA"] <- NA
}
# Combine how many times with yes or no column to get rid of the Non applicable (-3)
if("chrtbi_symptoms" %in% colnames(merged_data)){
  merged_data$chrtbi_symptoms <- paste(merged_data$chrtbi_subject_head_injury, merged_data$chrtbi_subject_symptoms)
  merged_data$chrtbi_symptoms[merged_data$chrtbi_symptoms=="No NA"] <- "No"
  merged_data$chrtbi_symptoms[merged_data$chrtbi_symptoms=="NA NA"] <- NA
}
# Combine how many times with yes or no column to get rid of the Non applicable (-3)
if("chrtbi_knockedout" %in% colnames(merged_data)){
  merged_data$chrtbi_knockedout <- paste(merged_data$chrtbi_subject_head_injury, merged_data$chrtbi_subject_knockedout)
  merged_data$chrtbi_knockedout[merged_data$chrtbi_knockedout=="No NA"] <- "No"
  merged_data$chrtbi_knockedout[merged_data$chrtbi_knockedout=="NA NA"] <- NA
}

#Remove unneeded columns
if("chrtbi_age_inj" %in% colnames(merged_data)){
  merged_data <- select(merged_data, -c(chrtbi_subject_head_injury, chrtbi_subject_age, chrtbi_subject_times,
                                        chrtbi_subject_symptoms, chrtbi_subject_knockedout))
}

## Include these new variables in the meta data

meta <- rbind(meta, data.frame(
  ElementName = 'chrtbi_age_inj',
  Datatype = 'Traumatic Brain Injury',
  Modality = 'Brain_injury',
  stringsAsFactors = FALSE
))

meta <- rbind(meta, data.frame(
  ElementName = 'chrtbi_ntimes',
  Datatype = 'Traumatic Brain Injury',
  Modality = 'Brain_injury',
  stringsAsFactors = FALSE
))

meta <- rbind(meta, data.frame(
  ElementName = 'chrtbi_symptoms',
  Datatype = 'Traumatic Brain Injury',
  Modality = 'Brain_injury',
  stringsAsFactors = FALSE
))

meta <- rbind(meta, data.frame(
  ElementName = 'chrtbi_knockedout',
  Datatype = 'Traumatic Brain Injury',
  Modality = 'Brain_injury',
  stringsAsFactors = FALSE
))

# If variable is about a medication dose, change all -3 and -300 to 0
# Loop through columns that have 'dose' in their names (case-insensitive)
for (col in names(merged_data)) {
  if (grepl("dose", col, ignore.case = TRUE)) {
    merged_data[[col]][merged_data[[col]] %in% c(-3, -300)] <- 0
  }
}
# Loop through columns that have 'med' in their names (case-insensitive)
for (col in names(merged_data)) {
  if (grepl("med", col, ignore.case = TRUE)) {
    merged_data[[col]][merged_data[[col]] %in% c(-3, -300)] <- 0
  }
}


# Remove all -3 and -300 when possible
merged_data[merged_data == -3 | merged_data == -300] <- NA
data_month2[data_month2 == -3 | data_month2 == -300] <- NA

# names(merged_data)[sapply(merged_data, function(x) any(x %in% c(-3,-300), na.rm=TRUE))] #To test which columns are still containing -3 and -300.


```



### Cognition

#### Standardizing using HC
```{r}

penncnb_vars <- grep("^cnb", names(merged_data), value = TRUE)

# Filter HC group
hc_data <- merged_data[merged_data$phenotype == "HC", ]

# Function to compute z-scores based on HC stats
z_score_hc <- function(x, varname) {
  m <- mean(hc_data[[varname]], na.rm = TRUE)
  s <- sd(hc_data[[varname]], na.rm = TRUE)
  return((x - m) / s)
}

# Apply z-scoring to all identified variables
for (var in penncnb_vars) {
  merged_data[[var]] <- z_score_hc(merged_data[[var]], var)
}
```

```{r}
# Filter HC group
hc_data_month2 <- data_month2[data_month2$phenotype == "HC", ]

# Function to compute z-scores based on HC stats
z_score_hc_month2 <- function(x, varname) {
  m <- mean(hc_data_month2[[varname]], na.rm = TRUE)
  s <- sd(hc_data_month2[[varname]], na.rm = TRUE)
  return((x - m) / s)
}

# Apply z-scoring to all identified variables
for (var in penncnb_vars) {
  data_month2[[var]] <- z_score_hc_month2(data_month2[[var]], var)
}
```


### Save preprocessed data

```{r}
#Save complete merged_data. NOTE: Change to merged_data_CT if you want to include MRI
write_csv(merged_data, '/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Data/merged_data.csv')
write_csv(meta, '/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Data/merged_meta.csv')
write_csv(data_month2, '/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Data/data_month2.csv')

```

# Demographic table full sample

```{r}
# Source the demographics function
source("/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Code/GIT/Scripts/Modules/Demographictable/Demographictable_function.R")

comparison <- 'phenotype'

comparison_labels <- c("CHR" = "CHR", "HC" = "CC")

modality_order <- c(
  "Demographics",
  "study_info",
  "Psychoticism",
  "Detachment",
  "Internalising",
  "Functioning",
  "Cognition"
)

# Generate summary table
table <- demographic_table(merged_data, dictionary_DIR, comparison, comparison_labels, modality_order=modality_order)

# Print table
print(table)

# Save in a word document
save_as_docx(table, path = "Dem_table_fullSample.docx")
```

# Demographic table final sample with comparison between discovery and test

First load in the labels from both discovery and test with subject ids. 
```{r}
discovery_labels <- read.csv("~/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Code/MultiCluster/discovery_labels.csv")
test_labels <- read.csv("~/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Code/MultiCluster/test_labels.csv")
```

Create new column with discovery vs test sample
```{r}
data_finalSample = merged_data

data_finalSample$sample <- ifelse(data_finalSample$src_subject_id %in% discovery_labels$src_subject_id, yes="Discovery", no="Test")

data_finalSample <- dplyr::filter(
  data_finalSample,
  src_subject_id %in% discovery_labels$src_subject_id |
    src_subject_id %in% test_labels$src_subject_id
)

```


```{r}
comparison <- 'sample'

table <- demographic_table(data_finalSample, dictionary_DIR, comparison, modality_order=modality_order)

print(table)
save_as_docx(table, path = "Dem_table_finalSample.docx")

```


# Demographic table labels
```{r}
# Merge data

Labels_data_disc <- merge(discovery_labels, merged_data, by='src_subject_id')
Labels_data_test <- merge(test_labels, merged_data, by='src_subject_id')
```

## Discovery

```{r}
# Set path to dictionary
dictionary_DIR_diff <- "/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Feature selection/Complete_dictionary_differences.xlsx"

comparisons <- c(
  "Psychoticism",
  "Detachment",
  "Internalising",
  "Functioning",
  "Cognition",
  "final"
)

tables <- setNames(
  lapply(comparisons, function(comp) {
    demographic_table(Labels_data_disc, dictionary_DIR, comparison = comp, , modality_order=modality_order)
  }),
  comparisons
)

# Put them all into one Word document
library(officer)
library(flextable)

doc <- read_docx()

for (comp in names(tables)) {
  doc <- doc %>%
    body_add_par(paste0("Demographics by ", comp), style = "heading 1") %>%
    body_add_flextable(tables[[comp]]) %>%
    body_add_par("", style = "Normal")
}

print(doc, target = "demographics_all_comparisons.docx")

```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(flextable)
library(officer)

summarize_differences <- function(df, dictionary_DIR_diff, comparisons,
                                 comparison_labels_map = list(),
                                 fdr_alpha = 0.05,
                                 top_n = NULL,
                                 keep = c("sig", "all")) {

  keep <- match.arg(keep)

  res_long <- map_dfr(comparisons, function(comp) {

    comp_labels <- comparison_labels_map[[comp]]

    gt <- demographic_table(
      df, dictionary_DIR_diff,
      comparison = comp,
      comparison_labels = comp_labels,
      return = "gtsummary",
      modality_order=modality_order
    )

    tb <- gt$table_body %>%
      # keep variable rows (labels + levels)
      filter(row_type %in% c("label", "level")) %>%
      mutate(
        comparison = comp,
        keep_main = !is.na(effect_size)   # only the main row where effect_size is stored
      ) %>%
      filter(keep_main) %>%
      transmute(
        comparison,
        variable,
        label,
        p = p.value,
        fdr = suppressWarnings(as.numeric(p.adjust(p.value, method = "fdr"))),
        effect_size = effect_size
      )

    if (!is.null(top_n)) {
      tb <- tb %>%
        arrange(fdr) %>%
        slice_head(n = top_n)
    }

    tb
  })

  if (keep == "sig") {
    res_used <- res_long %>%
      filter(!is.na(fdr)) %>%
      filter(fdr < fdr_alpha)
  } else {
    res_used <- res_long
  }

  compact <- res_used %>%
    mutate(
      cell = ifelse(
        is.na(fdr),
        NA_character_,
        paste0("ES=", effect_size, "; FDR=", format(round(fdr, 3), nsmall = 3))
      )
    ) %>%
    select(label, comparison, cell) %>%
    distinct() %>%
    pivot_wider(names_from = comparison, values_from = cell)

  list(
    long_all = res_long,
    long_used = res_used,
    compact = compact
  )
}

# ---- Word export (compact table styled similarly to your full tables) ----
export_compact_to_word <- function(compact_df, file = "differences_summary.docx",
                                  title = "Differences summary across comparisons") {

  ft <- flextable(compact_df) %>%
    theme_vanilla() %>%        # clean style close to gtsummary defaults
    bold(part = "header") %>%
    autofit() %>%
    align(align = "left", part = "all") %>%
    valign(valign = "top", part = "all") %>%
    set_table_properties(layout = "autofit")


  doc <- read_docx() %>%
    body_add_par(title, style = "heading 1") %>%
    body_add_flextable(ft) %>%
    body_add_par("", style = "Normal")

  print(doc, target = file)
  invisible(file)
}
```

```{r}
comparisons <- c("Psychoticism", "Detachment", "Internalising",
                 "Functioning", "Cognition", "final")

out <- summarize_differences(
  Labels_data_disc, dictionary_DIR_diff,
  comparisons = comparisons,
  fdr_alpha = 0.05,
  keep = "all"   # "sig" = only FDR < alpha; "all" = everything
)

# View
out$compact

# Export to Word
export_compact_to_word(out$compact, file = "differences_summary.docx")
```



```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tibble)

# For two different fill scales in one plot:
#install.packages("ggnewscale")
library(ggnewscale)

# ----------------------------
# Helpers
# ----------------------------
parse_fdr_num <- function(x) {
  if (is.null(x)) return(NA_real_)
  x <- as.character(x)
  x <- trimws(x)
  if (length(x) == 0 || is.na(x) || x == "") return(NA_real_)
  if (grepl("^<", x)) return(suppressWarnings(as.numeric(sub("^<\\s*", "", x))))
  if (grepl("^>", x)) return(suppressWarnings(as.numeric(sub("^>\\s*", "", x))))
  suppressWarnings(as.numeric(x))
}

# Decide variable type from raw df column
# - continuous: numeric with >2 unique non-missing values and not just {0,1}
# - binary: numeric with values subset of {0,1} OR factor/character with exactly 2 observed levels
# - categorical: factor/character with >=3 observed levels
detect_var_type <- function(x) {
  x0 <- x
  x0 <- x0[!is.na(x0)]

  if (length(x0) == 0) return(NA_character_)

  if (is.numeric(x0)) {
    u <- sort(unique(x0))
    if (length(u) <= 2 && all(u %in% c(0, 1))) return("binary")
    return("continuous")
  }

  # character/factor/logical treated as categorical/binary
  if (is.logical(x0)) return("binary")

  xch <- as.character(x0)
  u <- unique(xch)
  if (length(u) == 2) return("binary")
  if (length(u) >= 3) return("categorical")

  # single-level
  return(NA_character_)
}

# For binary variables, choose "positive" level for proportion
# rule: if levels include common positives -> use that; else use the 2nd level
choose_positive_level <- function(x) {
  xch <- as.character(x)
  u <- unique(xch[!is.na(xch)])
  if (length(u) < 2) return(NA_character_)

  # common positive labels
  positives <- c("1", "yes", "true", "positive", "present", "high", "case", "chr")
  ulow <- tolower(u)

  hit <- u[match(TRUE, ulow %in% positives, nomatch = 0)]
  if (length(hit) >= 1 && !is.na(hit[1])) return(hit[1])

  # otherwise: use the second level in sorted unique values for stability
  u_sorted <- sort(u)
  u_sorted[2]
}

# ----------------------------
# Build heatmap data (means + proportions + categorical grey, with FDR p-values)
# ----------------------------
build_mixed_heatmap_data <- function(df,
                                     dictionary_DIR,
                                     comparisons,
                                     comparison_labels_map = list(),
                                     include_overall = FALSE,
                                     scale_continuous_within_variable = TRUE,
                                     # for binary scale you typically DO NOT want z-scoring:
                                     scale_binary_within_variable = FALSE) {

  res <- purrr::map_dfr(comparisons, function(comp) {

    if (!comp %in% names(df)) {
      warning(paste0("Skipping comparison '", comp, "' because column not found in df."))
      return(tibble())
    }

    # Prepare comparison factor exactly like demographic_table does
    df2 <- df
    df2[[comp]] <- as.character(df2[[comp]])
    comp_labels <- comparison_labels_map[[comp]]
    if (!is.null(comp_labels) && length(comp_labels) > 0) {
      df2[[comp]] <- factor(df2[[comp]], levels = names(comp_labels), labels = comp_labels)
    } else {
      df2[[comp]] <- as.factor(df2[[comp]])
    }
    df2 <- df2[!is.na(df2[[comp]]), , drop = FALSE]

    grp_levels <- levels(df2[[comp]])
    if (length(grp_levels) == 0) grp_levels <- sort(unique(as.character(df2[[comp]])))

    # Get p.adjusted (FDR) from your demographic_table output (ensures identical testing pipeline)
    gt <- demographic_table(
      df, dictionary_DIR,
      comparison = comp,
      comparison_labels = comp_labels,
      return = "gtsummary"
    )

    tb <- gt$table_body
    p_tbl <- tb %>%
      filter(row_type == "label") %>%
      transmute(
        variable,
        fdr_text = if ("p.adjusted" %in% names(tb)) as.character(p.adjusted) else NA_character_,
        fdr_num  = if ("p.adjusted" %in% names(tb)) vapply(p.adjusted, parse_fdr_num, numeric(1)) else NA_real_
      )

    # Use labels from gtsummary (already de-duplicated in your demographic_table)
    var_labels <- tb %>%
      filter(row_type == "label") %>%
      transmute(variable, label = as.character(label))

    vars <- intersect(var_labels$variable, names(df2))

    if (length(vars) == 0) {
      warning(paste0("Skipping comparison '", comp, "' because none of the gtsummary variables exist as columns in df."))
      return(tibble())
    }

    # Build one row per (variable, group)
    out <- purrr::map_dfr(vars, function(v) {
      x <- df2[[v]]
      vt <- detect_var_type(x)
      if (is.na(vt)) return(tibble())

      # compute per-group value
      by <- df2[[comp]]

      if (vt == "continuous") {
        tmp <- tibble(
          variable = v,
          group = as.character(by),
          value = as.numeric(x)
        ) %>%
          group_by(variable, group) %>%
          summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
          mutate(var_type = "continuous")

      } else if (vt == "binary") {
        if (is.numeric(x)) {
          tmp <- tibble(
            variable = v,
            group = as.character(by),
            value = as.numeric(x)
          ) %>%
            group_by(variable, group) %>%
            summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%  # mean of 0/1 = proportion
            mutate(var_type = "binary")
        } else {
          pos <- choose_positive_level(x)
          tmp <- tibble(
            variable = v,
            group = as.character(by),
            value = as.character(x)
          ) %>%
            group_by(variable, group) %>%
            summarise(value = mean(value == pos, na.rm = TRUE), .groups = "drop") %>%
            mutate(var_type = "binary")
        }

      } else { # categorical (>=3 levels)
        # We still want tiles + p-values, but no mean/proportion color
        tmp <- tibble(
          variable = v,
          group = grp_levels
        ) %>%
          mutate(value = NA_real_, var_type = "categorical")
      }

      tmp
    })

    if (nrow(out) == 0) return(tibble())

    out %>%
      left_join(var_labels, by = "variable") %>%
      left_join(p_tbl, by = "variable") %>%
      mutate(
        comparison = comp,
        # group index (k) based on factor level order in df2
        k = match(group, grp_levels),
        x = paste0(comp, " | ", group)
      ) %>%
      select(comparison, x, k, variable, label, var_type, value, fdr_text, fdr_num)
  })

  if (nrow(res) == 0) {
    warning("No heatmap data produced.")
    return(res)
  }

  # Create plotting values:
  # - continuous: optionally z-score within variable
  # - binary: keep as proportion (0-1) unless explicitly scaled
  # - categorical: NA (will be grey)
  res <- res %>%
    group_by(label) %>%
    mutate(
      value_plot_cont = ifelse(var_type == "continuous",
                               if (scale_continuous_within_variable) as.numeric(scale(value)) else value,
                               NA_real_),
      value_plot_bin  = ifelse(var_type == "binary",
                               if (scale_binary_within_variable) as.numeric(scale(value)) else value,
                               NA_real_)
    ) %>%
    ungroup()

  res
}

# ----------------------------
# Spacers between modalities
# ----------------------------
add_modality_spacers <- function(heat_df, spacer_suffix = " ") {
  df <- heat_df %>%
    mutate(modality = trimws(sub("\\|.*$", "", as.character(x))))

  modalities <- unique(df$modality)
  if (length(modalities) <= 1) return(df)

  spacers <- purrr::map_dfr(modalities[-length(modalities)], function(m) {
    df %>%
      filter(modality == m) %>%
      distinct(label) %>%
      transmute(
        comparison = NA_character_,
        x = paste0(m, spacer_suffix),
        k = NA_integer_,
        variable = NA_character_,
        label = label,
        var_type = "spacer",
        value = NA_real_,
        fdr_text = NA_character_,
        fdr_num = NA_real_,
        value_plot_cont = NA_real_,
        value_plot_bin = NA_real_
      )
  })

  bind_rows(df, spacers)
}

order_x_with_spacers <- function(heat_df, spacer_suffix = " ") {
  df <- heat_df %>%
    mutate(
      x_chr = as.character(x),
      is_spacer = var_type == "spacer" | (is.na(value) & grepl(paste0(spacer_suffix, "$"), x_chr)),
      modality = trimws(sub("\\|.*$", "", x_chr))
    )

  base_x <- df %>% filter(!is_spacer) %>% distinct(x_chr, modality)
  modality_order <- base_x$modality[!duplicated(base_x$modality)]

  x_levels <- unlist(lapply(modality_order, function(m) {
    xs <- base_x$x_chr[base_x$modality == m]
    c(xs, paste0(m, spacer_suffix))
  }))
  x_levels <- x_levels[x_levels != paste0(modality_order[length(modality_order)], spacer_suffix)]

  df %>%
    mutate(x = factor(x_chr, levels = x_levels)) %>%
    select(-x_chr, -is_spacer, -modality)
}


```

```{r}
export_heatmap <- function(plot_obj, heat_df,
                           file = "heatmap_group_means.svg",
                           dpi = 300,
                           width = NULL, height = NULL,
                           units = "in") {

  n_rows <- dplyr::n_distinct(heat_df$label)
  n_cols <- dplyr::n_distinct(heat_df$x)

  # Heuristic sizing if not provided
  if (is.null(width))  width  <- max(12, n_cols * 0.6)
  if (is.null(height)) height <- max(8,  n_rows * 0.22)

  ggsave(
    filename = file,
    plot = plot_obj,
    width = width,
    height = height,
    units = units,
    dpi = dpi,
    limitsize = FALSE
  )

  message("Saved: ", file, " (", width, " x ", height, " ", units, ", dpi=", dpi, ")")
}

```

```{r}
comparisons <- c("Psychoticism","Detachment","Internalising","Functioning","Cognition","final")

heat_df <- build_mixed_heatmap_data(
  Labels_data_disc, dictionary_DIR_diff,
  comparisons = comparisons,
  comparison_labels_map = list(),
  include_overall = FALSE,
  scale_continuous_within_variable = TRUE,
  scale_binary_within_variable = FALSE
)

heat_df_spaced <- heat_df %>%
  add_modality_spacers() %>%
  order_x_with_spacers()


```

```{r}
add_x_positions_with_gaps <- function(df, gap_size = 0.25, step_size = 1) {
  # df$x must be a factor with levels in plotting order
  lv <- levels(df$x)
  if (is.null(lv)) stop("df$x must be a factor with levels set.")

  # your real columns look like "Modality | group"
  is_real <- grepl("\\|", lv)
  is_spacer <- !is_real

  pos <- numeric(length(lv))
  cur <- 0
  for (i in seq_along(lv)) {
    if (is_spacer[i]) {
      cur <- cur + gap_size
    } else {
      cur <- cur + step_size
    }
    pos[i] <- cur
  }

  map <- tibble::tibble(x = factor(lv, levels = lv), x_pos = pos, is_real = is_real)
  df %>% dplyr::left_join(map, by = "x")
}

heat_df_spaced <- heat_df_spaced %>%
  add_x_positions_with_gaps(gap_size = 0.20)  # <- make this smaller/larger
```


```{r}
plot_mixed_heatmap_split <- function(heat_df,
                                     dictionary_DIR,

                                     # overall (patchwork) title + size
                                     title = "Heatmaps by variable type",
                                     overall_title_size = 14,

                                     # x-axis title (optionally shown per panel) + size
                                     x_title = "Modality | group",
                                     x_title_size = 10,
                                     show_x_title_on = c("cat"), # any of c("cont","bin","cat")

                                     # panel subtitles (each plot title) + size (per panel)
                                     cont_subtitle = "Continuous (means)",
                                     bin_subtitle  = "Binary (proportions)",
                                     cat_subtitle  = "Categorical (neutral tiles; p-values shown)",
                                     cont_subtitle_size = 12,
                                     bin_subtitle_size  = 12,
                                     cat_subtitle_size  = 12,

                                     # axis text sizes
                                     x_text_size = 8,
                                     y_text_size = 8,

                                     # header + variable label sizes on y-axis (ggtext markdown)
                                     header_text_size = 13,
                                     var_text_size    = NULL,   # defaults to y_text_size

                                     # legend titles + sizes (per panel) and legend text sizes (per panel)
                                     legend_title_cont = "Mean",
                                     legend_title_bin  = "Prop",
                                     legend_title_size_cont = 11,
                                     legend_text_size_cont  = 10,
                                     legend_title_size_bin  = 11,
                                     legend_text_size_bin   = 10,

                                     # legend key sizing
                                     legend_key_height = grid::unit(0.5, "cm"),
                                     legend_key_width  = grid::unit(0.35, "cm"),

                                     # p-value annotation controls + size
                                     annotate_p = TRUE,
                                     p_text_size = 2.4,
                                     alpha_sig = 0.05,

                                     # colours
                                     categorical_grey = "grey85",

                                     # modality ordering for header blocks
                                     modality_order = c("Psychoticism",
                                                        "Detachment",
                                                        "Internalising",
                                                        "Functioning",
                                                        "Cognition",
                                                        "Demographics",
                                                        "Study_info")) {

  # ---- deps ----
  if (!requireNamespace("dplyr", quietly = TRUE)) stop("Need package: dplyr")
  if (!requireNamespace("ggplot2", quietly = TRUE)) stop("Need package: ggplot2")
  if (!requireNamespace("patchwork", quietly = TRUE)) stop("Need package: patchwork")
  if (!requireNamespace("shadowtext", quietly = TRUE)) stop("Need package: shadowtext")
  if (!requireNamespace("RColorBrewer", quietly = TRUE)) stop("Need package: RColorBrewer")
  if (!requireNamespace("readxl", quietly = TRUE)) stop("Need package: readxl")
  if (!requireNamespace("tidyr", quietly = TRUE)) stop("Need package: tidyr")
  if (!requireNamespace("tibble", quietly = TRUE)) stop("Need package: tibble")
  if (!requireNamespace("ggtext", quietly = TRUE)) stop("Need package: ggtext")
  if (!requireNamespace("grid", quietly = TRUE)) stop("Need package: grid")

  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(shadowtext)

  if (is.null(var_text_size)) var_text_size <- y_text_size
  show_x_title_on <- intersect(show_x_title_on, c("cont", "bin", "cat"))

  # ensure x is a factor (your ordering/spacers should already be baked into levels)
  if (!is.factor(heat_df$x)) heat_df <- heat_df %>% mutate(x = factor(x))

  # If you added numeric x positions (to shrink spacer gaps), we use them.
  use_xpos <- "x_pos" %in% names(heat_df)
  if (use_xpos && !is.numeric(heat_df$x_pos)) {
    stop("heat_df$x_pos exists but is not numeric. Recompute x_pos as numeric.")
  }

  # label x only for real columns "Modality | group" (hide spacer ticks)
  x_labeler <- function(lv) ifelse(grepl("\\|", lv), lv, "")

  # breaks/labels for numeric x axis (label only real columns; spacers unlabeled)
  if (use_xpos) {
    x_breaks <- heat_df %>% distinct(x, x_pos) %>% arrange(x_pos)
    x_breaks_real <- x_breaks %>% filter(grepl("\\|", as.character(x)))
    breaks_vec <- x_breaks_real$x_pos
    labels_vec <- as.character(x_breaks_real$x)
  }

  # ---- dictionary mapping: variable -> Modality + dict row + ElementName (short y label) ----
  dict <- readxl::read_excel(dictionary_DIR)
  req <- c("ElementName", "Modality", "Aliases")
  miss <- setdiff(req, names(dict))
  if (length(miss) > 0) stop("Dictionary is missing columns: ", paste(miss, collapse = ", "))

  dict <- dict %>%
    mutate(
      .dict_row   = dplyr::row_number(),
      ElementName = as.character(ElementName),
      Modality    = as.character(Modality),
      Aliases     = as.character(Aliases)
    )

  split_aliases <- function(x) {
    if (is.na(x) || !nzchar(x)) return(character(0))
    toks <- unlist(strsplit(x, "[;\\|,\\n\\r]+", perl = TRUE))
    toks <- trimws(toks)
    toks <- toks[nzchar(toks)]
    unique(toks)
  }

  element_tbl <- dict %>%
    transmute(key = ElementName, Modality, .dict_row, .prio = 1L) %>%
    filter(!is.na(key) & key != "")

  alias_tbl <- dict %>%
    transmute(Modality, .dict_row, alias_vec = lapply(Aliases, split_aliases)) %>%
    tidyr::unnest(alias_vec) %>%
    transmute(key = as.character(alias_vec), Modality, .dict_row, .prio = 2L) %>%
    filter(!is.na(key) & key != "")

  lookup_raw <- bind_rows(element_tbl, alias_tbl)

  vars_present <- heat_df %>%
    filter(!is.na(variable) & variable != "") %>%
    distinct(variable) %>%
    pull(variable) %>%
    as.character()

  matches <- tibble::tibble(variable = vars_present) %>%
    left_join(lookup_raw, by = c("variable" = "key"))

  # pick best match per variable: ElementName match preferred, then alias,
  # within that first occurrence in dict
  best <- matches %>%
    filter(!is.na(Modality)) %>%
    arrange(variable, .prio, .dict_row) %>%
    group_by(variable) %>%
    slice(1) %>%
    ungroup() %>%
    left_join(dict %>% transmute(.dict_row, ElementName), by = ".dict_row")

  var_map <- tibble::tibble(variable = vars_present) %>%
    left_join(best %>% select(variable, Modality, .dict_row, ElementName), by = "variable") %>%
    mutate(
      Modality      = ifelse(is.na(Modality) | Modality == "", "Other / Unmapped", Modality),
      .dict_row     = ifelse(is.na(.dict_row), Inf, .dict_row),
      element_short = ifelse(is.na(ElementName) | ElementName == "", variable, ElementName)
    ) %>%
    select(variable, Modality, .dict_row, element_short)

  df0 <- heat_df %>%
    left_join(var_map, by = "variable") %>%
    mutate(
      Modality      = ifelse(is.na(Modality) | Modality == "", "Other / Unmapped", Modality),
      .dict_row     = ifelse(is.na(.dict_row), Inf, .dict_row),
      element_short = ifelse(is.na(element_short) | element_short == "", variable, element_short)
    )

  # ---- build header rows + variable rows, with bold headers ----
  build_blocked_df <- function(df_type) {

    mods_in_panel <- df_type %>%
      filter(!is.na(variable) & variable != "") %>%
      distinct(Modality) %>%
      pull(Modality)

    mods_present <- c(modality_order, setdiff(mods_in_panel, modality_order))
    mods_present <- c(setdiff(mods_present, "Other / Unmapped"),
                      intersect(mods_present, "Other / Unmapped"))

    ord <- df_type %>%
      filter(!is.na(variable) & variable != "") %>%
      distinct(variable, label, element_short, Modality, .dict_row) %>%
      mutate(
        Modality = factor(Modality, levels = mods_present),
        .mod_rank = match(as.character(Modality), mods_present)
      ) %>%
      arrange(.mod_rank, .dict_row, element_short, variable)

    # x levels depend on whether we use factor-x or numeric-x (still keep factor for ordering)
    x_levels <- levels(df_type$x)
    k_levels <- sort(unique(df_type$k))

    header_rows <- ord %>%
      distinct(Modality) %>%
      mutate(
        y_id = paste0("##", as.character(Modality)),
        variable = NA_character_,
        label = as.character(Modality),
        y_disp = as.character(Modality),
        .dict_row = -Inf
      ) %>%
      tidyr::crossing(x = factor(x_levels, levels = x_levels)) %>%
      tidyr::crossing(k = k_levels) %>%
      mutate(
        comparison = df_type$comparison[1],
        var_type = df_type$var_type[1],
        value = NA_real_,
        fdr_text = NA_character_,
        fdr_num = NA_real_,
        value_plot_cont = NA_real_,
        value_plot_bin  = NA_real_,
        x_pos = if (use_xpos) NA_real_ else NULL
      )

    real_rows <- df_type %>%
      mutate(
        y_id   = paste0(as.character(Modality), "__", as.character(variable)),
        y_disp = element_short
      )

    # y order: headers then variables within each modality
    y_levels <- c()
    for (m in mods_present) {
      vars_m <- ord %>% filter(as.character(Modality) == m)
      if (nrow(vars_m) == 0) next
      y_levels <- c(y_levels,
                    paste0("##", m),
                    paste0(m, "__", vars_m$variable))
    }

    out <- bind_rows(real_rows, header_rows) %>%
      mutate(y_id = factor(y_id, levels = rev(unique(y_levels))))

    y_map <- out %>%
      distinct(y_id, y_disp) %>%
      mutate(y_id_chr = as.character(y_id)) %>%
      { setNames(.$y_disp, .$y_id_chr) }

    y_lab <- function(vals) {
      vals_chr <- as.character(vals)
      is_header <- startsWith(vals_chr, "##")
      raw <- unname(y_map[vals_chr])

      header_html <- paste0(
        "<span style='font-size:", header_text_size, "pt; font-weight:800;'><b>",
        raw,
        "</b></span>"
      )
      var_html <- paste0(
        "<span style='font-size:", var_text_size, "pt;'>&nbsp;&nbsp;&nbsp;",
        raw,
        "</span>"
      )

      ifelse(is_header, header_html, var_html)
    }

    list(df = out, y_labeler = y_lab)
  }

  df_cont0 <- df0 %>% filter(var_type == "continuous")
  df_bin0  <- df0 %>% filter(var_type == "binary")
  df_cat0  <- df0 %>% filter(var_type == "categorical")

  cont_pack <- build_blocked_df(df_cont0)
  bin_pack  <- build_blocked_df(df_bin0)
  cat_pack  <- build_blocked_df(df_cat0)

  df_cont <- cont_pack$df
  df_bin  <- bin_pack$df
  df_cat  <- cat_pack$df

  # if using numeric x, make sure header rows got x_pos too (join via x)
  if (use_xpos) {
    x_map_all <- df0 %>% distinct(x, x_pos)
    df_cont <- df_cont %>% left_join(x_map_all, by = "x", suffix = c("", ".map")) %>%
      mutate(x_pos = dplyr::coalesce(x_pos, x_pos.map)) %>% select(-x_pos.map)
    df_bin  <- df_bin  %>% left_join(x_map_all, by = "x", suffix = c("", ".map")) %>%
      mutate(x_pos = dplyr::coalesce(x_pos, x_pos.map)) %>% select(-x_pos.map)
    df_cat  <- df_cat  %>% left_join(x_map_all, by = "x", suffix = c("", ".map")) %>%
      mutate(x_pos = dplyr::coalesce(x_pos, x_pos.map)) %>% select(-x_pos.map)
  }

  n_cont <- max(1, dplyr::n_distinct(df_cont$y_id))
  n_bin  <- max(1, dplyr::n_distinct(df_bin$y_id))
  n_cat  <- max(1, dplyr::n_distinct(df_cat$y_id))

  # ---- annotations (stars on k==1, p text on k==2), skip headers ----
  base_ann <- df0 %>%
    filter(k == 1) %>%
    filter(!is.na(fdr_text) & fdr_text != "") %>%
    mutate(
      sig_stars = dplyr::case_when(
        grepl("^\\s*<\\s*0\\.001\\s*$", fdr_text) ~ "***",
        !is.na(fdr_num) & fdr_num < 0.001 ~ "***",
        !is.na(fdr_num) & fdr_num < 0.01  ~ "**",
        !is.na(fdr_num) & fdr_num < 0.05  ~ "*",
        TRUE ~ ""
      ),
      y_id = paste0(as.character(Modality), "__", as.character(variable))
    )

  map_k2 <- df0 %>%
    filter(k == 2) %>%
    mutate(y_id = paste0(as.character(Modality), "__", as.character(variable))) %>%
    distinct(comparison, y_id, x)

  ann_k1 <- base_ann %>%
    transmute(comparison, x, y_id, ann_label = sig_stars) %>%
    filter(ann_label != "")

  ann_k2 <- base_ann %>%
    select(comparison, y_id, fdr_text) %>%
    left_join(map_k2, by = c("comparison", "y_id")) %>%
    filter(!is.na(x)) %>%
    transmute(comparison, x, y_id, ann_label = fdr_text)

  ann <- bind_rows(ann_k1, ann_k2)

  # if numeric x, bring x_pos into ann for plotting
  if (use_xpos) {
    x_map_all <- df0 %>% distinct(x, x_pos)
    ann <- ann %>% left_join(x_map_all, by = "x")
  }

  add_p_layer <- function(p, df_section) {
    if (!annotate_p) return(p)

    ann_section <- ann %>%
      semi_join(df_section, by = c("x", "y_id")) %>%
      filter(!startsWith(as.character(y_id), "##"))

    p +
      shadowtext::geom_shadowtext(
        data = ann_section,
        aes(x = if (use_xpos) x_pos else x, y = y_id, label = ann_label),
        inherit.aes = FALSE,
        color = "white",
        bg.colour = "black",
        bg.r = 0.12,
        size = p_text_size
      )
  }

  # helper: x title per panel
  x_title_for <- function(panel) if (panel %in% show_x_title_on) x_title else NULL

  # ---- base themes ----
  base_theme_top <- theme_classic() +
    theme(
      axis.title.x = element_text(size = x_title_size),
      axis.title.y = element_blank(),
      axis.text.y  = ggtext::element_markdown(size = y_text_size),
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x  = element_blank(),

      legend.key.height = legend_key_height,
      legend.key.width  = legend_key_width
    )

  base_theme_bot <- theme_classic() +
    theme(
      axis.title.x = element_text(size = x_title_size),
      axis.title.y = element_blank(),
      axis.text.x  = element_text(angle = 45, hjust = 1, size = x_text_size),
      axis.text.y  = ggtext::element_markdown(size = y_text_size),

      legend.key.height = legend_key_height,
      legend.key.width  = legend_key_width
    )

  # ---- x scale (discrete vs continuous) ----
  x_scale_top <- function() {
    if (use_xpos) {
      scale_x_continuous(breaks = breaks_vec, labels = labels_vec, expand = c(0, 0))
    } else {
      scale_x_discrete(labels = x_labeler, drop = FALSE)
    }
  }
  x_scale_bot <- x_scale_top

  # ---- plots ----
  # NOTE: when using numeric x, geom_tile width should match the "step_size" used for real columns.
  # If your step_size is 1, width=1 looks good. If you change step_size, adjust width accordingly.
  tile_width <- if (use_xpos) 1 else 0.9

  p_cont <- ggplot(df_cont, aes(x = if (use_xpos) x_pos else x, y = y_id, fill = value_plot_cont)) +
    geom_tile(width = tile_width) +
    scale_fill_gradientn(colours = RColorBrewer::brewer.pal(9, "Blues"),
                         na.value = "white",
                         name = legend_title_cont) +
    x_scale_top() +
    scale_y_discrete(labels = cont_pack$y_labeler, drop = FALSE) +
    labs(title = cont_subtitle, x = x_title_for("cont"), y = NULL) +
    base_theme_top +
    theme(
      plot.title   = element_text(size = cont_subtitle_size),
      legend.title = element_text(size = legend_title_size_cont),
      legend.text  = element_text(size = legend_text_size_cont)
    )
  p_cont <- add_p_layer(p_cont, df_cont)

  p_bin <- ggplot(df_bin, aes(x = if (use_xpos) x_pos else x, y = y_id, fill = value_plot_bin)) +
    geom_tile(width = tile_width) +
    scale_fill_gradientn(colours = RColorBrewer::brewer.pal(9, "YlOrRd"),
                         limits = c(0, 1),
                         na.value = "white",
                         name = legend_title_bin) +
    x_scale_top() +
    scale_y_discrete(labels = bin_pack$y_labeler, drop = FALSE) +
    labs(title = bin_subtitle, x = x_title_for("bin"), y = NULL) +
    base_theme_top +
    theme(
      plot.title   = element_text(size = bin_subtitle_size),
      legend.title = element_text(size = legend_title_size_bin),
      legend.text  = element_text(size = legend_text_size_bin)
    )
  p_bin <- add_p_layer(p_bin, df_bin)

  # categorical: grey tiles only for non-header rows
  df_cat_tiles <- df_cat %>% filter(!startsWith(as.character(y_id), "##"))

  p_cat <- ggplot(df_cat_tiles, aes(x = if (use_xpos) x_pos else x, y = y_id)) +
    geom_tile(fill = categorical_grey, width = tile_width) +
    x_scale_bot() +
    scale_y_discrete(labels = cat_pack$y_labeler, drop = FALSE) +
    labs(title = cat_subtitle, x = x_title_for("cat"), y = NULL) +
    base_theme_bot +
    theme(
      plot.title = element_text(size = cat_subtitle_size)
    )
  p_cat <- add_p_layer(p_cat, df_cat_tiles)

  (p_cont / p_bin / p_cat) +
    patchwork::plot_layout(heights = c(n_cont, n_bin, n_cat)) +
    patchwork::plot_annotation(
      title = title,
      theme = theme(plot.title = element_text(size = overall_title_size))
    )
}
```



```{r}
map <- plot_mixed_heatmap_split(
  heat_df_spaced,
  dictionary_DIR = dictionary_DIR_diff,
  title = "My heatmap ordered by modality",
  overall_title_size = 20,

  # p-value annotation text
  p_text_size = 4,

  # axis text sizes
  x_text_size = 11,
  y_text_size = 11,

  # y-axis header/variable styling
  header_text_size = 13,
  var_text_size = 11,

  # subplot “subtitles” (panel titles) + sizes
  cont_subtitle = "Continuous (means)",
  bin_subtitle  = "Binary (proportions)",
  cat_subtitle  = "Categorical",
  cont_subtitle_size = 16,
  bin_subtitle_size  = 16,
  cat_subtitle_size  = 16,

  # x-axis title text + size
  x_title = "Modality | group",
  x_title_size = 14,
  show_x_title_on = c("cat"),

  # legend sizing (per panel)
  legend_title_size_cont = 12,
  legend_text_size_cont  = 11,
  legend_title_size_bin  = 12,
  legend_text_size_bin   = 11
)

export_heatmap(map, heat_df_spaced, file = "heatmap_group_means_ordered.svg",
               dpi = 300, height = 18, width = 14)

map

```


# Site and recruitment source
```{r}
# Packages
library(dplyr)
library(ggplot2)

# Ensure factors (optional but recommended)
merged_data <- merged_data %>%
  mutate(
    sites = as.factor(Site),
    chrrecruit = as.factor(chrrecruit)
  )

# Contingency table
tab <- table(merged_data$sites, merged_data$chrrecruit)
tab

# Chi-square test (overall association)
chi <- chisq.test(tab)
chi

# If expected counts are small, the chi-square approximation can be shaky:
chi$expected

# Option A: Monte Carlo p-value (good practical fallback)
chisq.test(tab, simulate.p.value = TRUE, B = 20000)

# Option B: Fisher's exact test (can be slow/too large for big tables)
# fisher.test(tab)
```

```{r}
site_recruit_dist <- ggplot(merged_data, aes(x = sites, fill = chrrecruit)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Site",
    y = "Proportion within site",
    fill = "Recruitment method",
    title = "Recruitment method distribution by site"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'site_recruitment_dist.pdf',
    plot = site_recruit_dist,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )
```

```{r}
plot_df <- merged_data %>%
  count(sites, chrrecruit, name = "n") %>%
  group_by(sites) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

site_recruit_plot <- ggplot(plot_df, aes(x = chrrecruit, y = sites, fill = prop)) +
  geom_tile() +
  scale_fill_viridis_c(labels = scales::percent_format()) +
  labs(
    x = "Recruitment method",
    y = "Site",
    fill = "Within-site %",
    title = "Within-site recruitment proportions"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'site_recruitment.pdf',
    plot = site_recruit_plot,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )
```

```{r}
library(tidyr)
library(seriation)

wide <- plot_df %>%
  select(sites, chrrecruit, prop) %>%
  pivot_wider(names_from = chrrecruit, values_from = prop, values_fill = 0)

mat <- as.matrix(wide[,-1])
rownames(mat) <- wide$sites

row_ord <- get_order(seriate(dist(mat), method = "OLO"))
col_ord <- get_order(seriate(dist(t(mat)), method = "OLO"))

plot_df2 <- plot_df %>%
  mutate(
    sites = factor(sites, levels = rownames(mat)[row_ord]),
    chrrecruit = factor(chrrecruit, levels = colnames(mat)[col_ord])
  )

site_recruit_plot2 <- ggplot(plot_df2, aes(chrrecruit, sites, fill = prop)) +
  geom_tile() +
  scale_fill_viridis_c(labels = scales::percent_format()) +
  labs(x = "Recruitment method", y = "Site", fill = "Within-site %") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'site_recruitment2.pdf',
    plot = site_recruit_plot2,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

site_recruit_plot2
```

# Site differences 
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(stringr)

subgroup_vars <- c("Psychoticism", "Detachment", "Internalising",
                   "Functioning", "Cognition", "final")

# Long format: one row per person x subgroup type
long_df <- Labels_data_disc %>%
  select(Site, all_of(subgroup_vars)) %>%
  mutate(
    Site = as.factor(Site),
    across(all_of(subgroup_vars), as.character)   # <- key line
  ) %>%
  pivot_longer(
    cols = all_of(subgroup_vars),
    names_to = "subgroup_type",
    values_to = "subgroup_label"
  ) %>%
  mutate(
    subgroup_type  = factor(subgroup_type, levels = subgroup_vars),
    subgroup_label = factor(subgroup_label)
  ) %>%
  filter(!is.na(Site), !is.na(subgroup_label))

# Helper: chi-square with Monte Carlo p-value + Cramer's V
one_test <- function(df) {
  tab <- table(df$Site, df$subgroup_label)

  # Chi-square with simulated p-value (robust when expected counts are small)
  chi <- suppressWarnings(chisq.test(tab, simulate.p.value = TRUE, B = 20000))

  # Cramer's V
  n <- sum(tab)
  r <- nrow(tab); c <- ncol(tab)
  V <- sqrt(unname(chi$statistic) / (n * (min(r - 1, c - 1))))

  tibble(
    statistic = unname(chi$statistic),
    df = unname(chi$parameter),
    p_value = unname(chi$p.value),
    cramers_v = V,
    n = n,
    n_sites = r,
    n_labels = c
  )
}

test_results <- long_df %>%
  group_by(subgroup_type) %>%
  group_modify(~ one_test(.x)) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adj)

test_results
```

```{r}
library(ggplot2)
library(forcats)
library(scales)

plot_df <- long_df %>%
  count(subgroup_type, Site, subgroup_label, name = "n") %>%
  group_by(subgroup_type, Site) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(plot_df, aes(x = Site, y = prop, fill = subgroup_label)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~ subgroup_type, ncol = 2) +
  labs(
    x = "Site",
    y = "Proportion within site",
    fill = "Subgroup label",
    title = "Subgroup label distribution by site (six subgroup definitions)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)

subgroup_vars <- c("Psychoticism", "Detachment", "Internalising",
                   "Functioning", "Cognition", "final")

long_df <- Labels_data_disc %>%
  select(Site, all_of(subgroup_vars)) %>%
  mutate(
    Site = as.factor(Site),
    across(all_of(subgroup_vars), as.character)
  ) %>%
  pivot_longer(all_of(subgroup_vars),
               names_to = "subgroup_type",
               values_to = "subgroup_label") %>%
  filter(!is.na(Site), !is.na(subgroup_label)) %>%
  mutate(
    subgroup_type = factor(subgroup_type, levels = subgroup_vars),
    subgroup_label = as.factor(subgroup_label)
  )

one_test <- function(df) {
  # drop unused levels inside this subgroup
  df <- df %>%
    mutate(
      Site = droplevels(Site),
      subgroup_label = droplevels(subgroup_label)
    )

  tab <- table(df$Site, df$subgroup_label)

  # remove any all-zero rows/cols (extra safety)
  tab <- tab[rowSums(tab) > 0, colSums(tab) > 0, drop = FALSE]

  # if degenerate after cleanup, return NA
  if (nrow(tab) < 2 || ncol(tab) < 2) {
    return(tibble(
      statistic = NA_real_, df = NA_integer_, p_value = NA_real_,
      cramers_v = NA_real_, n = sum(tab),
      n_sites = nrow(tab), n_labels = ncol(tab)
    ))
  }

  chi <- suppressWarnings(chisq.test(tab, simulate.p.value = TRUE, B = 20000))

  n <- sum(tab)
  r <- nrow(tab); c <- ncol(tab)
  V <- sqrt(unname(chi$statistic) / (n * (min(r - 1, c - 1))))

  tibble(
    statistic = unname(chi$statistic),
    df = unname(chi$parameter),
    p_value = unname(chi$p.value),
    cramers_v = V,
    n = n,
    n_sites = r,
    n_labels = c
  )
}

test_results <- long_df %>%
  group_by(subgroup_type) %>%
  group_modify(~ one_test(.x)) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adj)

test_results
```
```{r}
site_subgroups <- ggplot(plot_df, aes(x = subgroup_label, y = Site, fill = prop)) +
  geom_tile() +
  facet_wrap(~ subgroup_type, scales = "free_x") +
  scale_fill_viridis_c(labels = percent_format()) +
  labs(
    x = "Subgroup label",
    y = "Site",
    fill = "Within-site %",
    title = "Heatmap of subgroup label proportions within site"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'site_subgroup.pdf',
    plot = site_subgroups,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

site_subgroups
```

```{r}
facet_labs <- test_results %>%
  mutate(
    facet = paste0(
      subgroup_type,
      "\nBH p=", format.pval(p_adj, digits = 2, eps = 1e-4),
      " | V=", round(cramers_v, 2)
    )
  ) %>%
  select(subgroup_type, facet)

plot_df2 <- plot_df %>%
  left_join(facet_labs, by = "subgroup_type")

site_subgroups2 <- ggplot(plot_df2, aes(x = Site, y = prop, fill = subgroup_label)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~ facet, ncol = 2) +
  labs(
    x = "Site",
    y = "Proportion within site",
    fill = "Subgroup label",
    title = "Subgroup-by-site differences with BH-adjusted p-values and effect sizes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'site_subgroup2.pdf',
    plot = site_subgroups2,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

site_subgroups2
```

# Recruitment differences
```{r}
library(dplyr)
library(tidyr)

subgroup_vars <- c("Psychoticism", "Detachment", "Internalising",
                   "Functioning", "Cognition", "final")

df_long <- Labels_data_disc %>%
  select(Site, chrrecruit, all_of(subgroup_vars)) %>%
  mutate(
    Site = as.factor(Site),
    chrrecruit = as.factor(chrrecruit),
    across(all_of(subgroup_vars), as.character)   # avoid type-mix in pivot
  ) %>%
  pivot_longer(
    cols = all_of(subgroup_vars),
    names_to = "subgroup_type",
    values_to = "subgroup_label"
  ) %>%
  filter(!is.na(Site), !is.na(chrrecruit), !is.na(subgroup_label)) %>%
  mutate(
    subgroup_type  = factor(subgroup_type, levels = subgroup_vars),
    subgroup_label = factor(subgroup_label)
  )
```

```{r}
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)

subgroup_vars <- c("Psychoticism", "Detachment", "Internalising",
                   "Functioning", "Cognition", "final")

long_df_recruit <- Labels_data_disc %>%
  select(chrrecruit, all_of(subgroup_vars)) %>%
  mutate(
    chrrecruit = as.factor(chrrecruit),
    across(all_of(subgroup_vars), as.character)
  ) %>%
  pivot_longer(all_of(subgroup_vars),
               names_to = "subgroup_type",
               values_to = "subgroup_label") %>%
  filter(!is.na(chrrecruit), !is.na(subgroup_label)) %>%
  mutate(
    subgroup_type = factor(subgroup_type, levels = subgroup_vars),
    subgroup_label = as.factor(subgroup_label)
  )

one_test <- function(df) {
  # drop unused levels inside this subgroup
  df <- df %>%
    mutate(
      chrrecruit = droplevels(chrrecruit),
      subgroup_label = droplevels(subgroup_label)
    )

  tab <- table(df$chrrecruit, df$subgroup_label)

  # remove any all-zero rows/cols (extra safety)
  tab <- tab[rowSums(tab) > 0, colSums(tab) > 0, drop = FALSE]

  # if degenerate after cleanup, return NA
  if (nrow(tab) < 2 || ncol(tab) < 2) {
    return(tibble(
      statistic = NA_real_, df = NA_integer_, p_value = NA_real_,
      cramers_v = NA_real_, n = sum(tab),
      n_sites = nrow(tab), n_labels = ncol(tab)
    ))
  }

  chi <- suppressWarnings(chisq.test(tab, simulate.p.value = TRUE, B = 20000))

  n <- sum(tab)
  r <- nrow(tab); c <- ncol(tab)
  V <- sqrt(unname(chi$statistic) / (n * (min(r - 1, c - 1))))

  tibble(
    statistic = unname(chi$statistic),
    df = unname(chi$parameter),
    p_value = unname(chi$p.value),
    cramers_v = V,
    n = n,
    n_sites = r,
    n_labels = c
  )
}

test_results_recruit <- long_df_recruit %>%
  group_by(subgroup_type) %>%
  group_modify(~ one_test(.x)) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_adj)

test_results_recruit
```

```{r}
plot_df_recruit <- long_df_recruit %>%
  count(subgroup_type, chrrecruit, subgroup_label, name = "n") %>%
  group_by(subgroup_type, chrrecruit) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(plot_df_recruit, aes(x = chrrecruit, y = prop, fill = subgroup_label)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~ subgroup_type, ncol = 2) +
  labs(
    x = "Recruitment source",
    y = "Proportion within recruitment source",
    fill = "Subgroup label",
    title = "Subgroup label distribution by recruitment (six subgroup definitions)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
recruitment_subgroups <- ggplot(plot_df_recruit, aes(x = subgroup_label, y = chrrecruit, fill = prop)) +
  geom_tile() +
  facet_wrap(~ subgroup_type, scales = "free_x") +
  scale_fill_viridis_c(labels = percent_format()) +
  labs(
    x = "Subgroup label",
    y = "Recruitment source",
    fill = "Within-recruitment %",
    title = "Heatmap of subgroup label proportions within recruitment source"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggsave(
    filename = 'recruitment_subgroup.pdf',
    plot = recruitment_subgroups,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

recruitment_subgroups
```

```{r}
facet_labs <- test_results_recruit %>%
  mutate(
    facet = paste0(
      subgroup_type,
      "\nBH p=", format.pval(p_adj, digits = 2, eps = 1e-4),
      " | V=", round(cramers_v, 2)
    )
  ) %>%
  select(subgroup_type, facet)

plot_df2 <- plot_df_recruit %>%
  left_join(facet_labs, by = "subgroup_type")

recruitment_subgroups2 <- ggplot(plot_df2, aes(x = chrrecruit, y = prop, fill = subgroup_label)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~ facet, ncol = 2) +
  labs(
    x = "Recruitment source",
    y = "Proportion within recruitment",
    fill = "Subgroup label",
    title = "Subgroup-by-recruitment differences with BH-adjusted p-values and effect sizes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = 'recruitment_subgroup2.pdf',
    plot = recruitment_subgroups2,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

recruitment_subgroups2
```

# Site+recruitment differences subgroups

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(nnet)      # multinom
library(broom)

subgroup_vars <- c("Psychoticism","Detachment","Internalising","Functioning","Cognition","final")

df_long <- Labels_data_disc %>%
  select(Site, chrrecruit, all_of(subgroup_vars)) %>%
  mutate(
    Site = factor(Site),
    chrrecruit = factor(chrrecruit),
    across(all_of(subgroup_vars), as.character)
  ) %>%
  pivot_longer(all_of(subgroup_vars), names_to="subgroup_type", values_to="subgroup_label") %>%
  filter(!is.na(Site), !is.na(chrrecruit), !is.na(subgroup_label)) %>%
  mutate(
    subgroup_type = factor(subgroup_type, levels = subgroup_vars),
    subgroup_label = factor(subgroup_label)
  )

fit_one_type <- function(d){
  d <- d %>%
    mutate(
      Site = droplevels(Site),
      chrrecruit = droplevels(chrrecruit),
      subgroup_label = droplevels(subgroup_label)
    )

  if (nlevels(d$subgroup_label) < 2) return(NULL)

  # Models
  m0      <- multinom(subgroup_label ~ 1,         data = d, trace = FALSE)
  m_site  <- multinom(subgroup_label ~ Site,      data = d, trace = FALSE)
  m_recr  <- multinom(subgroup_label ~ chrrecruit,data = d, trace = FALSE)
  m_full  <- multinom(subgroup_label ~ Site + chrrecruit, data = d, trace = FALSE)

  # Unadjusted effects (vs intercept-only)
  site_unadj   <- anova(m0, m_site, test = "Chisq")
  recr_unadj   <- anova(m0, m_recr, test = "Chisq")

  # Adjusted effects (incremental)
  # Recruitment adjusted for Site: (Site) -> (Site + Recruit)
  recr_adj_site <- anova(m_site, m_full, test = "Chisq")

  # Site adjusted for Recruitment: (Recruit) -> (Site + Recruit)
  site_adj_recr <- anova(m_recr, m_full, test = "Chisq")

  tibble(
    lr_site_unadjusted = site_unadj$`LR stat.`[2],
    p_site_unadjusted  = site_unadj$`Pr(Chi)`[2],

    lr_recruit_unadjusted = recr_unadj$`LR stat.`[2],
    p_recruit_unadjusted  = recr_unadj$`Pr(Chi)`[2],

    lr_recruit_adj_for_site = recr_adj_site$`LR stat.`[2],
    p_recruit_adj_for_site  = recr_adj_site$`Pr(Chi)`[2],

    lr_site_adj_for_recruit = site_adj_recr$`LR stat.`[2],
    p_site_adj_for_recruit  = site_adj_recr$`Pr(Chi)`[2]
  )
}

model_results <- df_long %>%
  group_by(subgroup_type) %>%
  group_modify(~{
    out <- fit_one_type(.x)
    if (is.null(out)) tibble(
      lr_site_unadjusted = NA_real_, p_site_unadjusted = NA_real_,
      lr_recruit_unadjusted = NA_real_, p_recruit_unadjusted = NA_real_,
      lr_recruit_adj_for_site = NA_real_, p_recruit_adj_for_site = NA_real_,
      lr_site_adj_for_recruit = NA_real_, p_site_adj_for_recruit = NA_real_
    ) else out
  }) %>%
  ungroup() %>%
  mutate(
    # BH-adjust each “question family” across the 6 subgroup schemes
    p_site_unadjusted_BH        = p.adjust(p_site_unadjusted, method = "BH"),
    p_recruit_unadjusted_BH     = p.adjust(p_recruit_unadjusted, method = "BH"),
    p_recruit_adj_for_site_BH   = p.adjust(p_recruit_adj_for_site, method = "BH"),
    p_site_adj_for_recruit_BH   = p.adjust(p_site_adj_for_recruit, method = "BH")
  ) %>%
  rename(subgroup_scheme = subgroup_type)

model_results

```

```{r}
library(marginaleffects)
library(dplyr)

plot_one_type <- function(d, type_name){
  d <- d %>%
    mutate(Site = droplevels(Site),
           chrrecruit = droplevels(chrrecruit),
           subgroup_label = droplevels(subgroup_label))

  m_site <- nnet::multinom(subgroup_label ~ Site, data=d, trace=FALSE)
  m_full <- nnet::multinom(subgroup_label ~ Site + chrrecruit, data=d, trace=FALSE)

  p1 <- predictions(m_site, newdata = datagrid(Site = levels(d$Site))) %>%
    rename(outcome_level = group) %>%  # <- outcome category column
    mutate(model = "Unadjusted (Site only)")

  p2 <- avg_predictions(m_full, variables = "Site") %>%
    rename(outcome_level = group) %>%
    mutate(model = "Adjusted (Site + recruitment)")

  bind_rows(p1, p2) %>%
    mutate(subgroup_type = type_name)
}

pred_df <- df_long %>%
  group_split(subgroup_type) %>%
  purrr::map_dfr(~ plot_one_type(.x, unique(.x$subgroup_type))) %>%
  filter(!is.na(outcome_level), outcome_level != "NA") %>%
  mutate(outcome_level = factor(outcome_level))

ggplot(pred_df, aes(x = Site, y = estimate, color = model)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0, position = position_dodge(width = 0.5)) +
  facet_grid(subgroup_type ~ outcome_level, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
plot_df <- df_long %>%
  count(subgroup_type, chrrecruit, Site, subgroup_label, name="n") %>%
  group_by(subgroup_type, chrrecruit, Site) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

site_recruitment_subgroups <- ggplot(plot_df, aes(x = Site, y = prop, fill = subgroup_label)) +
  geom_col(position="fill") +
  facet_grid(subgroup_type ~ chrrecruit, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x="Site", y="Proportion within site", fill="Subgroup label") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))

ggsave(
    filename = 'site_recruitment_subgroup.pdf',
    plot = site_recruitment_subgroups,
    width = 20,
    height = 20,
    dpi = 300,
    limitsize = FALSE
  )

site_recruitment_subgroups
```
```{r}
one_chi <- function(d){
  d <- d %>% mutate(Site = droplevels(Site), subgroup_label = droplevels(subgroup_label))
  tab <- table(d$Site, d$subgroup_label)
  tab <- tab[rowSums(tab)>0, colSums(tab)>0, drop=FALSE]
  if (nrow(tab) < 2 || ncol(tab) < 2) return(tibble(p_value=NA))
  chi <- suppressWarnings(chisq.test(tab, simulate.p.value=TRUE, B=20000))
  tibble(p_value = unname(chi$p.value))
}

strat_tests <- df_long %>%
  group_by(subgroup_type, chrrecruit) %>%
  group_modify(~ one_chi(.x)) %>%
  ungroup() %>%
  group_by(subgroup_type) %>%
  mutate(p_adj = p.adjust(p_value, method="BH")) %>%
  ungroup()

strat_tests
```


## Test sample

```{r}
# Set path to dictionary
dictionary_DIR_diff <- "/Users/jentemeijer/Library/CloudStorage/OneDrive-Personal/OneDrive/PhD/Feature selection/Complete_dictionary_differences.xlsx"

comparisons <- c(
  "Psychoticism",
  "Detachment",
  "Internalising",
  "Functioning",
  "Cognition",
  "final"
)

tables_test <- setNames(
  lapply(comparisons, function(comp) {
    demographic_table(Labels_data_test, dictionary_DIR_diff, comparison = comp, , modality_order=modality_order)
  }),
  comparisons
)

# Put them all into one Word document

doc_test <- read_docx()

for (comp in names(tables_test)) {
  doc_test <- doc_test %>%
    body_add_par(paste0("Demographics by ", comp), style = "heading 1") %>%
    body_add_flextable(tables_test[[comp]]) %>%
    body_add_par("", style = "Normal")
}

print(doc_test, target = "demographics_all_comparisons_testsample.docx")

```

```{r}
comparisons <- c("Psychoticism", "Detachment", "Internalising",
                 "Functioning", "Cognition", "final")

out_test <- summarize_differences(
  Labels_data_test, dictionary_DIR_diff,
  comparisons = comparisons,
  fdr_alpha = 0.05,
  keep = "all"   # "sig" = only FDR < alpha; "all" = everything
)

# View
out_test$compact

# Export to Word
export_compact_to_word(out_test$compact, file = "differences_summary_testsample.docx")
```

